---
title: "Exploring Potential Influencers of Math Student Performance"
author: "Lauren Duvall, Shikyna William, Thomas Linden"
date: "`r Sys.Date()`"
lang: en
output: 
  html_document:
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(dplyr))    # Data manipulation
suppressMessages(library(ggplot2))  # Plotting
suppressMessages(library(plotly))  # Plotting
suppressMessages(library(readr))    # Reading the dataset
suppressMessages(library(knitr))    # For clean tables
```

[Kaggle Link](https://www.kaggle.com/datasets/adilshamim8/math-students/data)

## Introduction

This dataset

___

## Load Packages
```{r}
library(dplyr)
library(plotly)
```

___

## Data Cleaning
```{r}
raw <- read.csv("Math-Students.csv")
#Rename cols, convert factors for summary, transmute to drop old columns
formatted <- raw %>% transmute(
  school = school,
  sex = sex,
  age = age,
  address = address,
  fam_size = famsize,
  parental_stat = Pstatus,
  mom_edu = Medu,
  dad_edu = Fedu,
  mom_job = Mjob,
  dad_job = Fjob,
  attend_reason = reason,
  guardian = guardian,
  travel_time = traveltime,
  study_time = studytime,
  failures = failures,
  school_support = schoolsup,
  family_support = famsup,
  extra_paid_classes = paid,
  activities = activities,
  nursery_school = nursery,
  pursue_higher_edu = higher,
  internet_use = internet,
  romantic = romantic,
  family_relationship = famrel,
  free_time = freetime,
  go_out_w_friends = goout,
  workday_alcohol = Dalc,
  weekend_alcohol = Walc,
  health = health,
  absences = absences,
  period1_grade = G1,
  period2_grade = G2,
  period3_grade = G3
)
summary(formatted)
```

___

## Exploratory Data Analysis
```{r}
#Make correlation table
cor_df <- formatted %>% select(where(is.numeric)) %>% cor(use = "complete.obs") %>% 
  as.table() %>% as.data.frame()

colnames(cor_df) <- c("Var1", "Var2", "Correlation")

cor_df %>% ggplot(aes(Var1, Var2, fill = Correlation)) + 
  geom_tile(color = 'white') + 
  scale_fill_gradient2(low = '#067bc2', mid = 'white', high = '#fe4a49', 
                       midpoint = 0, limit = c(-1,1), name = 'Correlation') +
  labs(title = "Correlation Heatmap", x = '', y = '') + 
  theme_minimal() + 
  coord_fixed() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

**Intuitive relationships:** 



**Some relationships to look into more:**



**Unintuitive relationships:** 



**Features not in the correlation heatmap:** 

___

## Shikyna - School & Demographics

## Shikyna - Family Background

## Academic Support & Study Habits

## Activities and Social Life

## Lauren - Health & Lifestyle

## Lauren - Academic Performance

___

## Conclusions

### Limitations:

### Future Steps:

___

## References:

> Shamim, A. (n.d.). Math-Students Performance Data [Data set]. Kaggle. https://www.kaggle.com/datasets/adilshamim8/math-students/data

> P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira (Eds.), Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008), pp. 5-12, Porto, Portugal, April 2008, EUROSIS, ISBN 978-9077381-39-7.