---
title: "Shikyna stuff"
author: "RIntro_William"
date: "2025-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(dplyr))    # Data manipulation
suppressMessages(library(ggplot2))  # Plotting
suppressMessages(library(plotly))   # Plotting
suppressMessages(library(readr))    # Reading the dataset
suppressMessages(library(knitr))
```

## From a gazillion datasets, why this one?

- Clean, Structured and small:
Easy to work with, making it suitable for both beginners and intermediate learners. Even a 7 year old would be able to look at it and make some comment on it.

- Focused on Math Performance:
Unlike other data sets, this one is specified on math student performance (like a zoom in), hence more insights and you can go in more depth.

- Rich Demographic Features which demonstrate equity and policy insights:
Includes data like gender, parental education, lunch type, and test preparation.

- Equity and Policy Insights:
Enables exploration of how social and economic factors influence education.

------------------------------------------------------------------------

## Students and alcohol? Romantic relationships? Is that even legal?

Valid question.

- Reality Check:
Just because something is not legal doesn’t mean it doesn’t happen. We're trying to understand how real-life behaviors (like drinking or dating) might affect school performance.

- Cultural Context:
The dataset is from Portugal, where norms can be a little different. For example, some families may introduce teens to alcohol in small, supervised amounts, especially during meals.

- Insight, Not Endorsement:
Including this data doesn’t mean anyone’s encouraging it. It’s about spotting trends. Like—“Do students in relationships do better or worse in math?” or “Is there a pattern between alcohol use and test scores?”

- Helping, Not Policing: The goal is to help educators and policymakers understand what’s really going on so they can support students better, not to get anyone in trouble.

------------------------------------------------------------------------

```{r echo=FALSE}
library(dplyr)
library(plotly)
library(DT)
``` 


```{r echo=FALSE}
raw <- read.csv("Math-Students.csv")
#Rename cols, convert to factors, add any necessary levels, transmute to drop old columns
formatted <- raw %>% transmute(
  school = as.factor(school),
  sex = as.factor(sex),
  age = age,
  address = as.factor(address),
  fam_size = as.factor(famsize),
  parental_stat = as.factor(Pstatus),
  mom_edu = factor(Medu, levels = 0:4, ordered = TRUE),
  dad_edu = factor(Fedu, levels = 0:4, ordered = TRUE),
  mom_job = as.factor(Mjob),
  dad_job = as.factor(Fjob),
  attend_reason = as.factor(reason),
  guardian = as.factor(guardian),
  travel_time = traveltime,
  study_time = studytime,
  failures = failures,
  school_support = as.factor(schoolsup),
  family_support = as.factor(famsup),
  extra_paid_classes = as.factor(paid),
  activities = as.factor(activities),
  nursery_school = as.factor(nursery),
  pursue_higher_edu = as.factor(higher),
  internet_use = as.factor(internet),
  romantic = as.factor(romantic),
  family_relationship = factor(famrel, levels = 1:5, ordered = TRUE),
  free_time = factor(freetime, levels = 1:5, ordered = TRUE),
  go_out_w_friends = factor(goout, levels = 1:5, ordered = TRUE),
  workday_alcohol = factor(Dalc, levels = 1:5, ordered = TRUE),
  weekend_alcohol = factor(Walc, levels = 1:5, ordered = TRUE),
  health = factor(health, levels = 1:5, ordered = TRUE),
  absences = absences,
  period1_grade = G1,
  period2_grade = G2,
  period3_grade = G3
)
```

```{r}
datatable(formatted, 
          options = list(pageLength = 10,
                         scrollX = TRUE, 
                         autoWidth = TRUE, 
                         columnDefs = list(list(className = 'dt-center', targets = "_all"))), 
          rownames = FALSE)
sum(is.na(formatted))
```
#### 🏫 School & Demographics

- Do students from different schools (GP vs MS) show significant differences in final performance (G3)? INSIGNIFICANT

GP = Gabriel Pereira
MS = Mousinho da Silveira

```{r}
# Prepare the data
plot_data <- formatted %>%
  select(school, period3_grade) %>%   # Select relevant columns (school and final grade G3)
  group_by(school) %>%     # Group by school (GP vs MS)
  summarise(
    mean_period3 = mean(period3_grade, na.rm = TRUE),
    median_period3 = median(period3_grade, na.rm = TRUE),
    sd_period3 = sd(period3_grade, na.rm = TRUE),
    count = n()
  )

# View summary statistics
plot_data

#plot
formatted %>%
  ggplot(aes(x = school, y = period3_grade, fill = school)) +
  geom_violin(trim = FALSE, alpha = 0.8, color = NA) +
  geom_boxplot(width = 0.1, fill = "white", outlier.color = "white", outlier.shape = 16) +
  scale_fill_manual(values = c("GP" = "red", "MS" = "blue")) +
  labs(
    title = "Final Grade Distribution by School",
    subtitle = "Comparison of GP vs MS students' performance",
    x = "School",
    y = "Final Grade (Period 3)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text = element_text(color = "gray20"),
    panel.grid.major.y = element_line(color = "gray85"),
    legend.position = "top"
  )
```




- Is there an interaction between romantic relationships and gender on the final grade?

```{r}
formatted %>%
  ggplot(aes(x = romantic, y = period3_grade, fill = sex)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.6, outlier.shape = 16, alpha = 0.7) +
  scale_fill_manual(values = c("F" = "#FF69B4", "M" = "#1E90FF")) +
  labs(
    title = "Performance by Romantic Status and Gender",
    subtitle = "Do relationships affect boys and girls differently?",
    x = "Romantic Relationship",
    y = "Final Grade (Period 3)",
    fill = "Gender"
  ) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white", color = "black", position = position_dodge(0.8)) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text = element_text(color = "gray20"),
    panel.grid.major.y = element_line(color = "gray85"),
    legend.position = "top"
  )
```




#### 👪 Family Background


- Does having family support at home (famsup) help students perform better in school?


```{r}
formatted %>%
  ggplot(aes(x = as.factor(period3_grade), fill = family_support)) +
  geom_bar(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.9) +
  scale_fill_manual(values = c("yes" = "red", "no" = "blue")) +
  labs(
    title = "Impact of Family Support on Student Performance",
    subtitle = "Distribution of Final Math Grades Based on Family Support",
    x = "Final Grade",
    y = "Number of Students",
    fill = "Family Support at Home"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text = element_text(color = "gray20"),
    panel.grid.major.y = element_line(color = "gray85"),
    legend.position = "top"
  )
```


