---
title: "Investigating Contributing Factors of Obesity in Latin American Countries"
author: "Lauren Duvall, Shikyna William, Thomas Linden"
date: "`r Sys.Date()`"
lang: en
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(dplyr))    # Data manipulation
suppressMessages(library(ggplot2))  # Plotting
suppressMessages(library(plotly))  # Plotting
suppressMessages(library(readr))    # Reading the dataset
suppressMessages(library(knitr))    # For clean tables
```

[Link](https://www.kaggle.com/datasets/adeniranstephen/obesity-prediction-dataset)

## Introduction

This dataset contains mostly synthetically generated data, with some gathered from users via a web platform. The dataset attempts to estimate obesity levels across various health and lifestyle features from individuals in Mexico, Peru, and Columbia.

Although this dataset's primary purpose is to train ML models on obesity data for predictive analytics, the scope of this project is descriptive.

## Load Packages
```{r}
library(dplyr)
library(plotly)
```

## Data Cleaning
```{r}
raw <- read.csv("ObesityDataSet_raw_and_data_sinthetic.csv")
#Rename cols, convert factors for summary, transmute to drop old columns
formatted <- raw %>% transmute(
  gender = as.factor(Gender),
  age = Age,
  height = Height,
  weight = Weight * 2.20462, #convert to lbs
  family_hist_overweight = as.factor(family_history_with_overweight),
  high_calorie_foods_frequently = as.factor(FAVC),
  vegetable_frequency = FCVC,
  daily_main_meals = NCP,
  snack_frequency = factor(CAEC, levels = c("Never", "Sometimes", "Frequently", "Always"), ordered = TRUE),
  smoker = as.factor(SMOKE),
  daily_water_intake = CH2O,
  monitors_calories = as.factor(SCC),
  phys_activity_frequency = FAF,
  time_using_tech = TUE,
  alcohol_frequency = factor(CALC, levels = c("no", "Sometimes", "Frequently", "Always"), ordered = TRUE),
  main_transportation = as.factor(MTRANS),
  obesity_level = factor(NObeyesdad, levels = c(
    "Insufficient_Weight", "Normal_Weight", 
    "Overweight_Level_I", "Overweight_Level_II", 
    "Obesity_Type_I", "Obesity_Type_II", "Obesity_Type_III"), ordered = TRUE)
)
summary(formatted)
```

## Exploratory Data Analysis
```{r}
#Make correlation table
cor_df <- formatted %>% select(where(is.numeric)) %>% cor(use = "complete.obs") %>% 
  as.table() %>% as.data.frame()

colnames(cor_df) <- c("Var1", "Var2", "Correlation")

cor_df %>% ggplot(aes(Var1, Var2, fill = Correlation)) + 
  geom_tile(color = 'white') + 
  scale_fill_gradient2(low = '#067bc2', mid = 'white', high = '#fe4a49', 
                       midpoint = 0, limit = c(-1,1), name = 'Correlation') +
  labs(title = "Correlation Heatmap", x = '', y = '') + 
  theme_minimal() + 
  coord_fixed() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Some intuitive relationships:
- Height and Weight, Age and Weight

Some relationships to look into more:
- Phys Activity, Water Intake, Weight
- Phys Activity, Height, and Weight

Some unintuitive relationships:
- Vegetables and Weight

Not in Cor:
- gender, hist, calie,snack, smoker, monitors, alchohol, transport, obesity level

## Weight vs. Age by Obesity Levels
```{r}
formatted %>% 
  ggplot(aes(x = age, y = weight, color = obesity_level)) + geom_point(alpha = .5) + geom_smooth() + 
  guides(color = guide_legend(reverse = TRUE)) + 
  geom_vline(xintercept = 41, linetype = 'dashed') +
  theme_bw() +
  labs(y = 'Weight', x = 'Age', color = 'Obesity Level', title = 'Age vs. Weight by Obesity Level')
```
Although the dataset does not track life span or mortality, there are less records at higher ages for unhealthy obesity levels. This indicates either lifestyle change, or shortened life expectancy of those categories. 

Obesity Type III is an extreme case of weight gain and may have been oversampled when 77% of the data was synthesized. Alternatively, its possible these individuals had a medical wake-up call and the sharp decline for this group could indicate drastic lifestyle changes or survivorship.



## Conclusions

### Limitations:

### Future Steps:

## References:
> Mendoza Palechor, F., & De la Hoz Manotas, A. (2019). Dataset for estimation of obesity levels based on eating habits and physical condition in individuals from Colombia, Peru and Mexico. Data in Brief, 25, 104344. https://doi.org/10.1016/j.dib.2019.104344

> Mendoza Palechor, F., & De la Hoz Manotas, A. (2019). Estimation of obesity levels based on eating habits and physical condition [Data set]. UCI Machine Learning Repository. https://archive.ics.uci.edu/dataset/544/estimation+of+obesity+levels+based+on+eating+habits+and+physical+condition

> Stephen, A. (n.d.). Obesity Prediction Dataset [Data set]. Kaggle. https://www.kaggle.com/datasets/adeniranstephen/obesity-prediction-dataset