---
title: "Investigating Contributing Factors of Obesity in Latin American Countries"
author: "Lauren Duvall, Shikyna William, Thomas Linden"
date: "`r Sys.Date()`"
lang: en
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(dplyr))    # Data manipulation
suppressMessages(library(ggplot2))  # Plotting
suppressMessages(library(readr))    # Reading the dataset
suppressMessages(library(knitr))    # For clean tables
```

[Link](https://www.kaggle.com/datasets/adeniranstephen/obesity-prediction-dataset)

## Introduction

This dataset contains mostly synthetically generated data, with some gathered from users via a web platform. The dataset attempts to estimate obesity levels across various health and lifestyle features from individuals in Mexico, Peru, and Columbia.

Although this dataset's primary purpose is to train ML models on obesity data for predictive analytics, the scope of this project is descriptive.

## Load Packages
```{r}
library(dplyr)
```

## Data Cleaning
```{r}
raw_data <- read.csv("ObesityDataSet_raw_and_data_sinthetic.csv")
```

## Summary of Variables

## Exploratory Data Analysis

### Obesity Level Distribution

### Physical Activity vs Obesity Level

### Snacking Habits vs Obesity Level

## Conclusions

### Limitations:

### Future Steps:

## References:
> Mendoza Palechor, F., & De la Hoz Manotas, A. (2019). Dataset for estimation of obesity levels based on eating habits and physical condition in individuals from Colombia, Peru and Mexico. Data in Brief, 25, 104344. https://doi.org/10.1016/j.dib.2019.104344

> Mendoza Palechor, F., & De la Hoz Manotas, A. (2019). Estimation of obesity levels based on eating habits and physical condition [Data set]. UCI Machine Learning Repository. https://archive.ics.uci.edu/dataset/544/estimation+of+obesity+levels+based+on+eating+habits+and+physical+condition

> Stephen, A. (n.d.). Obesity Prediction Dataset [Data set]. Kaggle. https://www.kaggle.com/datasets/adeniranstephen/obesity-prediction-dataset